---
title: "Capstone Analysis"
author: "Polly Haun"
date: "2/3/2022"
output: html_document
---

#Load packages and libraries 
#Import NHANES 2017-2018 data using nhanesA package and create data frame

```{r setup, include=FALSE}
pacman::p_load(survival, ggfortify, survminer, readxl, ggplot2, 
               tidyverse, lmtest, table1, odds.n.ends, nhanesA, openxlsx, car, 
               nnet, blorr)

nhanesTables("EXAM", 2017)
nhanesTables("DEMO", 2017)
nhanesTables("QUESTIONNAIRE", 2017)
nhanesTableVars("QUESTIONNAIRE", "IMQ_J")
nhanesTableVars("DEMO", "DEMO_J")
demo <- nhanes("DEMO_J") #demographics
imq <- nhanes("IMQ_J") #immunizations
hiq <- nhanes("HIQ_J") #health insurance
inc <- nhanes("INQ_J") #income
huq <- nhanes("HUQ_J") #hospital utilization and access to health care

#create data frame
nhanes1 <- merge(demo,imq)
nhanes2 <- merge(hiq, inc)
nhanes3 <- merge(nhanes2, huq)
nhanes <- merge(nhanes1, nhanes3)

options(digits=4)

#rename variables
#nhanes <- nhanesTranslate("DEMO_J", "RIAGENDR", data=demo)
#nhanes <- nhanesTranslate("DEMO_J", "RIDRETH3", data=demo)
#nhanes <- nhanesTranslate("DEMO_J", "DMDEDUC3", data=demo)
#nhanes <- nhanesTranslate("DEMO_J", "DMDEDUC2", data=demo)
#nhanes <- nhanesTranslate("IMQ_J", "IMQ060")
#summary(nhanes)

#create smaller data frame with only people 18-26 years old
nhanes <- nhanes %>%
  filter(RIDAGEYR %in% (18:26))
```

#Data wrangling
#Rename and recode variables
```{r}

#change column names 
nhanes <- nhanes %>%
  rename(gender = "RIAGENDR",
         age = "RIDAGEYR",
        race = "RIDRETH3",
          education1 = "DMDEDUC2",
          education2 = "DMDEDUC3",
         householdIncome = "INDHHIN2",
          hpvFemale = "IMQ060",
          hpvMale = "IMQ070",
          hpvDoses = "IMQ100",
        ageAtVax = "IMQ090",
        place = "HUQ030")

#recode gender
nhanes<-nhanes %>%
  mutate(gender=case_when(gender %in% "1" ~ 0,
                          gender %in% "2" ~ 1))
nhanes$gender<-factor(nhanes$gender, levels=c(0:1), labels=c("Male", "Female"))

#recode race
nhanes<-nhanes%>%
  mutate(race=case_when(race %in% "3" ~ 0,
                         race %in% c("1", "2") ~ 1,
                         race %in% "4" ~ 2, 
                        race %in% "6" ~ 3,
                        race %in% "7" ~ 4))
nhanes$race<-factor(nhanes$race, levels=c(0:4), labels=c("Non-Hispanic White", "Hispanic or Latino", "Black or African American", "Asian", "Other"))
table(nhanes$race)

#recode binary vaccine variables into single variable  
nhanes<-nhanes %>%
  mutate(hpvShot=case_when(hpvFemale %in% "2" ~ 0,
                           hpvMale %in% "2" ~ 0,
                           hpvFemale %in% "1" ~ 1,
                           hpvMale %in% "1" ~ 1))
nhanes$hpvShot<-factor(nhanes$hpvShot, levels=c(0:1), labels=c("No", "Yes"))
table(nhanes$hpvShot)

#code missing values for hpvDoses
nhanes$hpvDoses[nhanes$hpvDoses == 9] <- NA 
nhanes$ageAtVax[nhanes$ageAtVax == 999] <- NA

#recode education into a single variable
table(nhanes$DMDEDUC2)
table(nhanes$DMDEDUC3)
nhanes<-nhanes %>%
  mutate(edu=case_when(education1 %in% c("1", "2") ~ 0,
                           education1 %in% "3" ~ 1,
                           education1 %in% "4" ~ 2,
                           education1 %in% "5" ~ 3,
                          education2 %in% c("9", "10", "11", "12") ~ 0,
                          education2 %in% c("13","14") ~ 1,
                          education2 %in% "15" ~ 2))
nhanes$edu<-factor(nhanes$edu, levels=c(0:3), labels=c("Less than high school", "High school or GED", "Some college", "College or above"))
table(nhanes$edu)

#recode insurance into a single variable
nhanes<-nhanes %>%
  mutate(insur=case_when(HIQ031D %in% "17" ~ 1, 
                         HIQ031F %in% "19" ~ 2, 
                         HIQ031H %in% "21" ~ 2, 
                         HIQ031I %in% "21" ~ 2,
                         HIQ031J %in% "21" ~ 2,
                         HIQ011 %in% "1" ~ 0,
                         HIQ011 %in% "2" ~ 3))
table(nhanes$insur)
nhanes$insur<-factor(nhanes$insur, levels=c(0:3), labels=c("Private", "Medicaid", "Other Insurance", "None"))

#recode routine place for healthcare
nhanes<-nhanes %>%
  mutate(place=case_when(place %in% c("1", "3") ~ 0,
                                place %in% "2" ~ 1))
nhanes$place<-factor(nhanes$place, levels=c(0:1), labels=c("Yes", "No"))


#need to exclude values lower than 9 for ageAtVax
#there is probably a more parsimonious way to do this but this is what's working now
nhanes$ageAtVax <- na_if(nhanes$ageAtVax, "2")
nhanes$ageAtVax <- na_if(nhanes$ageAtVax, "7")
nhanes$ageAtVax <- na_if(nhanes$ageAtVax, "8")
table(nhanes$ageAtVax)

#filter to exclude HPV shots where the value is missing (n=129)
nhanes <- nhanes %>%
  filter(hpvShot %in% c("Yes", "No"))

nhanes <- nhanes[!is.na(nhanes$edu),]
nhanes <- nhanes[!is.na(nhanes$insur),]

```

#create Table1
```{r}

nhanes$age <- as.numeric(nhanes$age) 
nhanes$ageAtVax <- as.numeric(nhanes$ageAtVax)
nhanes$hpvDoses <- factor(nhanes$hpvDoses)  

label(nhanes$age) <-  "Age" 
label(nhanes$race) <- "Race/Ethnicity"
label(nhanes$edu) <- "Education"
label(nhanes$insur) <- "Insurance"
label(nhanes$place) <- "Regular Place for Healthcare"
label(nhanes$gender) <- "Gender"

pvalue <- function(x, ...) {
    # Construct vectors of data y, and groups (strata) g
    y <- unlist(x)
    g <- factor(rep(1:length(x), times=sapply(x, length)))
    if (is.numeric(y)) {
        # For numeric variables, perform a standard 2-sample t-test
        p <- t.test(y ~ g)$p.value
    } else {
        # For categorical variables, perform a chi-squared test of independence
        p <- chisq.test(table(y, g))$p.value
    }
    # Format the p-value, using an HTML entity for the less-than sign.
    # The initial empty string places the output on the line below the variable label.
    c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))}

table(nhanes$ageAtVax)
class(nhanes$ageAtVax)

table1 <- table1(~age + race + edu + insur + place + gender|hpvShot, overall= FALSE, rowlabelhead="Variable", extra.col=list("P-value"=pvalue), data=nhanes)
table1

table1T <- table1(~age + race + edu + insur + place + gender|hpvShot, overall= "Total", rowlabelhead="Variable", data=nhanes)

#openxlsx::write.xlsx(table1, "C:/Users/polly/Box/Class Files/Capstone/Table1.xlsx")
#openxlsx::write.xlsx(table1T, "C:/Users/polly/Box/Class Files/Capstone/Table1T.xlsx")
```


#logistic regression of gender and HPV vaccination status (yes/no) 
#Analyze "Place" variable to see if it is an effect modifier or confounder of the relationship

```{r}
unadjModel1 <- glm(hpvShot ~ gender, data=nhanes, family="binomial")
summary(unadjModel1)
odds.n.ends(unadjModel1)

#chi-squared = 61.77, p < 0.001, df = 1
#model sensitivity = 55.4%, model specificity = 73.1%
#crude OR = 3.39 (95% CI 2.48-4.62)
#Among the study population, people who had received the HPV vaccine were 3.38 times more likely to be female than male. 

unadjModel2 <- glm(hpvShot ~ gender + place, data=nhanes, family="binomial")
summary(unadjModel2)
odds.n.ends(unadjModel2)

#chi-squared = 101.22, p < 0.001, df = 2
#model sensitivity = 87.0%, model specificity = 40.2%
#adjusted OR = 2.92 (95% CI 2.55-4.03)

#include interaction term 
unadjModel3 <- glm(hpvShot ~ gender + place + hpvShot*place, data=nhanes, family="binomial")
summary(unadjModel3)
odds.n.ends(unadjModel3)
#interaction term is not significant, so no effect modification

#stratum specific values for place
placeYes <- nhanes %>%
  filter(place == "Yes")
placeNo <- nhanes %>%
  filter(place == "No")

ssYes <- glm(hpvShot ~ gender, data=placeYes, family="binomial")
summary(ssYes)
odds.n.ends(ssYes)
#stratum specific OR for Yes place for health care = 2.79 (95% CI 1.89-4.16)

ssNo <- glm(hpvShot ~ gender, data=placeNo, family="binomial")
summary(ssNo)
odds.n.ends(ssNo)
#stratum specific OR for No place for health care = 3.18 (95% CI 1.83-5.58)

```

#adjusted model

```{r}
adjModel1 <- glm(hpvShot ~ gender + place + age + race + edu + insur, data=nhanes, family="binomial")
summary(adjModel1)
odds.n.ends(adjModel1)
#chi-squared = 115.91, p < 0.001, df = 13
#model sensitivity = 55.1%, model specificity = 76.4%
#adjusted OR = 2.78 (95% CI 1.99-3.90)
```

#Test Model 1 for assumptions

```{r}
#Box-Tidwell Test
#linearity for age, which is the only continuous predictor
nhanes <- nhanes %>%
  mutate(age.times.logage = age * log(age)) #create term to test linearity

boxTidwellAge <- glm(hpvShot ~ age + age.times.logage, data=nhanes, family="binomial") #Box Tidwell technique, test the assumption of linearity

summary(boxTidwellAge)
#findings are not significant, so the assumption of linearity is met 

```

```{r}
#Cook's Distance
plot(adjModel1, which=4, id.n=5, col="red") 
#influential observations are: 83, 225, 228, 377, 695

#set cut off point to 0.010
cutoff <- 0.010
obs_no <- as.data.frame(cooks.distance(adjModel1)) %>%
  mutate(obs_no=row_number()) %>%
  filter(`cooks.distance(adjModel1)` > cutoff)

#Let's exclude the values shown in the Cook's D plot, and see how the models compare.

model1.modex <- update(adjModel1,subset=c(-obs_no$obs_no))

#compare coefficients between models with and without influential observations, #caveat model number of observations is not the same
compareCoefs(adjModel1, model1.modex)
summary(model1.modex)
odds.n.ends(model1.modex)

#values for the model excluding observation 377
#chi-squared = 115.67, p < 0.001, df = 13
#model sensitivity = 54.8%, model specificity = 75.6%
#adjusted OR = 2.76 (95% CI 1.98-3.88)
#model is not signficantly different when excluding observation 377, so it will remain in the data frame

```

```{r}
#multicollinearity
vif(adjModel1)
#values are all <2, so multicollinearity is not present

#Hosmer Lemeshow test for goodness of fit
blr_test_hosmer_lemeshow(adjModel1)
#p value is >0.05 so model is not a bad fit
```

#Model 2
#Create a Table 1 for this model 
```{r}
#filter to only include values where the HPV dose is known 
nhanesDose <- nhanes %>%
  filter(hpvDoses %in% c(1, 2, 3))

#mutate hpvDoses variable so that it is binary
#1 or 2 doses = 0; 3 doses (up to date) = 1
nhanesDose<-nhanesDose %>%
  mutate(hpvDoses=case_when(hpvDoses %in% c("1","2") ~ 0,
                          hpvDoses %in% "3" ~ 1))
nhanesDose$hpvDoses<-factor(nhanesDose$hpvDoses, levels=c(0:1), labels=c("1 or 2 Doses", "3 Doses"))

pvalue <- function(x, ...) {
    # Construct vectors of data y, and groups (strata) g
    y <- unlist(x)
    g <- factor(rep(1:length(x), times=sapply(x, length)))
    if (is.numeric(y)) {
        # For numeric variables, perform a standard 2-sample t-test
        p <- t.test(y ~ g)$p.value
    } else {
        # For categorical variables, perform a chi-squared test of independence
        p <- chisq.test(table(y, g))$p.value
    }
    # Format the p-value, using an HTML entity for the less-than sign.
    # The initial empty string places the output on the line below the variable label.
    c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))}

table2 <- table1(~age + race + edu + insur+ place
                    + gender |hpvDoses, overall= F,
      rowlabelhead="Variable", extra.col=list("P-value"=pvalue), data=nhanesDose)
table2T <- table1(~age + race + edu + insur+ place
                    + gender |hpvDoses, overall= "Total",
      rowlabelhead="Variable", data=nhanesDose)
#openxlsx::write.xlsx(table2, "C:/Users/polly/Box/Class Files/Capstone/Table2.xlsx")
#openxlsx::write.xlsx(table2T, "C:/Users/polly/Box/Class Files/Capstone/Table2T.xlsx")

```


```{r}
unadjModel4 <- glm(hpvDoses ~ gender, data=nhanesDose, family="binomial")
summary(unadjModel4)
odds.n.ends(unadjModel4)

#chi-squared = 6.84, p < 0.01, df = 1
#model sensitivity = 79.1%, model specificity = 35.3%
#crude OR = 2.07 (95% CI 1.20-3.62)
#Among the study population, people who had completed the HPV vaccine series were 2.07 times more likely to be female than male. 

#include interaction term 
unadjModel5 <- glm(hpvDoses ~ gender + place + hpvDoses*place, data=nhanesDose, family="binomial")
summary(unadjModel5)
odds.n.ends(unadjModel5)
#interaction term is not significant

adjModel2 <- glm(hpvDoses ~ gender + place + age + race + edu + insur, data=nhanesDose, family="binomial")
summary(adjModel2)
odds.n.ends(adjModel2)
#chi-squared = 23.60, p < 0.05, df = 13
#model sensitivity = 74.3%, model specificity = 48.7%
#adjusted OR = 2.10 (95% CI 1.16-3.87)
```
#Test Model 2 for assumptions
```{r}
#Box-Tidwell Test
#linearity for age, which is the only continuous predictor
nhanesDose <- nhanesDose %>%
  mutate(age.times.logage = age * log(age)) #create term to test linearity

boxTidwellAge2 <- glm(hpvDoses ~ age + age.times.logage, data=nhanesDose, family="binomial") #Box Tidwell technique, test the assumption of linearity

summary(boxTidwellAge2)
#findings are not significant (p>0.05), so the assumption of linearity is met

#Cook's Distance
plot(adjModel2, which=4, id.n=5, col="red") 
#influential observations are: 19, 46, 128, 143, 254

#set cut off point to 0.015
cutoff <- 0.015
obs_no <- as.data.frame(cooks.distance(adjModel2)) %>%
  mutate(obs_no=row_number()) %>%
  filter(`cooks.distance(adjModel2)` > cutoff)

#Let's exclude the values shown in the Cook's D plot, and see how the models compare.

model2.modex <- update(adjModel2,subset=c(-obs_no$obs_no))

#compare coefficients between models with and without influential observations, #caveat model number of observations is not the same
compareCoefs(adjModel2, model2.modex)
summary(model2.modex)
odds.n.ends(model2.modex)
#Chi-squared increases slightly without obs 19
#Because it does not signifcantly change the model, I will not exclude the obs

#multicollinearity
vif(adjModel2)
#no values are >2 so multicollinearity is not present

#Hosmer Lemeshow test for goodness of fit
blr_test_hosmer_lemeshow(adjModel2)
#p is not significant (p>0.05) indicating that it is not a bad fit
```

